---
title: 'Red Wine Quality'
author: 'Tyler Borras'
output:
  html_document:
    fig_caption: yes
    keep_md: yes
    theme: sandstone
    toc: no
    toc_float: False
---

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library(ggplot2)
library(dplyr)
library(purrr)
library(tidyr)
library(ggcorrplot)

theme_set(theme_light())
```

```{r echo=FALSE}
#eliminate "#" from output
knitr::opts_chunk$set(comment = NA)

```

```{r echo=FALSE, Load_the_Data}
df <- read.csv('wineQualityReds.csv') # Cortez et al., 2009

```
## Introduction
>This data set contains physicochemical and sensory data for the red variant of 
Portuguese "Vinho Verde" wine. The dataset contains 11 independent variables 
and one dependent variable. The dependent variable is the median of the 
subjective "quality" of the wine as gauged by at least three experts.

## Univariate Plots Section

> This section will begin with summary statistics for the data and then the 
univariate plots.

```{r echo=FALSE, Summary}
glimpse(df)
summary(df)
```
> There are twelve columns of data and the index column which will be dropped. 
all columns are of type double except the dependent variable "quality", which is 
an integer. The features are measured using the following:

#### Input variables (based on physicochemical tests):
1. fixed acidity (tartaric acid - g / dm^3)
2. volatile acidity (acetic acid - g / dm^3)
3. citric acid (g / dm^3)
4. residual sugar (g / dm^3)
5. chlorides (sodium chloride - g / dm^3
6. free sulfur dioxide (mg / dm^3)
7. total sulfur dioxide (mg / dm^3)
8. density (g / cm^3)
9. pH
10. sulphates (potassium sulphate - g / dm3)
11. alcohol (% by volume)
* Output variable (based on sensory data): 
12. quality (score between 0 and 10)

(Cortez et al., 2009)

```{r echo=FALSE}
#Drop ther index column
drop_col1 <- c('X')
df <- df[ , !(names(df) %in% drop_col1)]

```


```{r echo=FALSE}
#print historgrams of all independent variables
drop_col2 <- c('quality')
print_df <- df
print_df <- print_df[ , !(names(print_df) %in% drop_col2)]

print_df %>%
  keep(is.numeric) %>% 
  gather() %>% 
  ggplot(aes(value)) +
    facet_wrap(~ key, scales = "free") +
    geom_histogram(fill = 'darkred',color = 'black', bins = 12)
```

```{r echo = FALSE}
#A function to print summaries of variables
p_sum <- function(p_var){
  print(summary(df[p_var]))
}

```

> A Quick view of the independent variables to get abetter idea of the 
distributions.

```{r echo=FALSE, Univariate_Plots}

c <- ggplot(df, aes(quality))
c + geom_bar( fill = 'darkred',color = 'black', size = 1)

```

```{r echo = FALSE}
p_sum('quality')

```

> The distribution of the quality appears to be normal, with a range of three to 
eight and a median of 6. There are very few samples for 3, 4, or 8.

```{r echo=FALSE}
c <- ggplot(df, aes(alcohol))
c + geom_area( stat = "bin",fill = 'darkred', color = 'black', size = 1, bins = 20)
```

```{r echo=FALSE}
p_sum('alcohol')

```

> Alcohol is skewed right and peaks around 9.5 with another smaller peak at 11.
The minimum of 8.4, maximum of 14.9, and median of 10.2 are all suprising 
because most wines, that I have seen, have very little 
variance in alcohol content.

```{r echo=FALSE}
c <- ggplot(df, aes(citric.acid ))
c + geom_histogram(binwidth = 0.025, fill = 'darkred', color = 'black')
```

```{r echo=FALSE}
p_sum('citric.acid')

```

> A closer look at citric acid reveals peaks in descending order at 0, 0.25, 
and 0.50. The first quartile is at 0.09 due to a large number of wines with 
0 citric acid.

```{r echo=FALSE}
c <- ggplot(df, aes(residual.sugar ))
c + geom_histogram(binwidth = 0.5, fill = 'darkred', color = 'black')
```

```{r echo=FALSE}
p_sum('residual.sugar')

```

> Residual sugar is highly skewed to the right with a median of 2.2 and a range 
of 0.9 to 15.5. Most of the wines fall into a very small range of sugar, the 
IQR is 1.9-2.6. I don't think this is going to be a good predictor due to the 
lack of variance.

```{r echo=FALSE}
c <- ggplot(df, aes(total.sulfur.dioxide ))
c + geom_density(kernel = "gaussian", fill = 'darkred', color = 'black', size = 1)

```

```{r echo=FALSE}
p_sum('total.sulfur.dioxide')

```

> Total sulfer dioxide is also highly right skewed with a median of 38 and a 
range of 6 to 289.

```{r echo = FALSE}
d <- ggplot(df, aes(free.sulfur.dioxide ))
d + geom_density(kernel = "gaussian", fill = 'darkred', color = 'black', size = 1)
```
```{r echo=FALSE}
p_sum('free.sulfur.dioxide')

```

> Free sulfer dioxide has a similar distribution to total sulfur dioxide, with 
a lower range of 1 to 72 with a median of 14.

```{r echo = FALSE}
d <- ggplot(df, aes(fixed.acidity))
d + geom_histogram( fill = 'darkred', color = 'black', size = 1, bins = 40)
```
```{r echo=FALSE}
p_sum('fixed.acidity')

```

> Fixed acidity is normally distributed and skewed right

```{r}
d <- ggplot(df, aes(volatile.acidity))
d + geom_histogram( fill = 'darkred', color = 'black', size = 1, bins = 40)
```

```{r echo=FALSE}
p_sum('volatile.acidity')

```

> Volatile acidity has a similar distribution to fixed acidity with a range of 
0.12 to 1.58 versus the range of 4.6 to 15.9 of fixed acidity

```{r echo = FALSE}
d <- ggplot(df, aes(pH))
d + geom_density(kernel = "gaussian", fill = 'darkred', color = 'black')

```
```{r echo = FALSE}
p_sum('pH')

```

> pH is normally distributed with a mean and median of 3.31 

# Univariate Analysis

### What is the structure of your dataset?
> There are eleven independent variables and one dependent variable. All 
variables are of type double except the dependent variable "quality", which is 
an integer. There are 1,599 samples. Quality is normally distributed with a 
range of 3 to 8 and a median of 6.

### What is/are the main feature(s) of interest in your dataset?
>Quality is the dependent variable so it is the main feature of interest.

### What other features in the dataset do you think will help support your 
### investigation into your feature(s) of interest?
>  If I had to speculate at this point in the analysis, I would say alcohol, 
and all of the acidity based variables including citric acid will be the most 
important supporting features.  


### Did you create any new variables from existing variables in the dataset?
> I have not created any variables at this point of the investigation.

### Of the features you investigated, were there any unusual distributions? \
> I consider citric acid to be unusual because so many samples have zero
citric acid. It seems unusual that wine from the same region made with the same
process would be void of a substance, while other wines contain a small amount.
Alcohol was also strange due to its large variance, I think it may be a good 
predictor.


# Bivariate Plots Section

```{r echo = FALSE}
g <- ggplot(df, aes(citric.acid))
g + geom_density(aes(fill=factor(quality)), alpha=0.8) 
```

> I started with citric acid because I thought it was the most unusual 
distribution of the univariate plots. This graph shows that low quality wines, 
those with quality of 3 or 4, generally have low, or zero, citric acid content. 
The wines with  quality of 5 or 6 generally have a moderate citric acid content. 
High quality wines, with aquality of 7 or 8, generally have the highest citric acid 
content. Citric acid definitely has a correlation with quality.

```{r echo=FALSE, Bivariate_Plots}
h <- ggplot(df, aes(citric.acid, pH))
h + geom_jitter(color = 'darkred') +
geom_smooth(method = lm, color = 'black', fill = 'green')
```

> Taking a closer look at citric acid reveals an inverse relationship with pH.
This was expected as a lower pH indicates higher acidity. Next, I would like to see 
the relationship between pH and quality.

```{r echo = FALSE}
g <- ggplot(df, aes(pH))
g + geom_density(aes(fill=factor(quality)), alpha=0.8)
```

> While not as distinct as the citric acid correlation to quality, pH does 
appear to have an inverse correlation with quality which confirms the findings
of the two previous graphs.

```{r echo=FALSE}
d <- ggplot(df, aes(factor(quality), pH, fill = factor(quality)))
d + geom_violin() + 
  geom_boxplot(width = 0.3)
```

> A closer look at pH reveals that wines with a quality of 3 have the highest 
median pH and a very small range. The median and the interquartile range trend
lower as the quality increases. I would like a closer look at the other acidic 
features to determine if they contribute to the quality as well.

```{r echo = FALSE}
d <- ggplot(df, aes(factor(quality), volatile.acidity, fill = factor(quality)))
d + geom_violin() + 
  geom_boxplot(width = 0.3)
```

>Both the mean and the IQR for volatile acidity decreases as the quality 
increases. While citric acid increases, volatile acidity decreases with quality,
possibly explaining why the trend for pH is more subtle.

```{r echo = FALSE}
d <- ggplot(df, aes(factor(quality), fixed.acidity, fill = factor(quality)))
d + geom_violin() + 
  geom_boxplot(width = 0.3)
```

> Fixed acidity does not appear to have any significant affect on quality.

```{r echo = FALSE}
d <- ggplot(df, aes(factor(quality), alcohol, fill = factor(quality)))
d + geom_violin() + 
  geom_boxplot(width = 0.2)
```

> The next interesting featurefrom the univariate analysis was alcohol. Alcohol
content in wines with a quality of 3, 4, or 5 tends to be low, while wines with 
a quality of 6, 7, or 8 generally have higher alcohol and a distinct upward 
trend. The wines with a quality of 5 have a large range, and actually have a 
lower median and IQR than the lower quality wines. It appears as though there is
some other factor other than alcohol in lower quality wines that affects the 
quality. High quality wines, 7 and 8, have a significantly higher median and IQR
than the other qualities.

```{r echo=FALSE}
h <- ggplot(df, aes(log10(alcohol), log10(residual.sugar)))
h + geom_jitter(color = 'darkred') +
geom_smooth(method = lm, color = 'black', fill = 'green')
```

> I thought sugar may play a part in the alcohol content, but despite their 
direct relationship in the fermentation process, there appears to be very 
little, if any, correlation between alcohol and residual sugar.

```{r echo = FALSE}
g <- ggplot(df, aes(residual.sugar))
g + geom_density(aes(fill=factor(quality)), alpha=0.8)  
    
```

>Residual sugar does not vary significantly for different quality ratings. This
confirms the observations made in the univariate section.

```{r echo = FALSE}
d <- ggplot(df, aes(factor(quality), sulphates, fill = factor(quality)))
d + geom_violin() + 
  geom_boxplot(width = 0.3)
```

>Higher levels of sulphates are present in the higher quality wines, with the 
median and IQR increasing with quality. Another interesting observation is the 
small range of sulphates for wines with quality 3 or 8, and a large range for 
wines in the middle.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the 
### investigation. How did the feature(s) of interest vary with other features 
### in the dataset?

> Citric acid, volatile acidity, sulphates, pH, and alcohol all had a 
correlation with quality.  Low, or zero, citric acid levels were present in 
low-quality wines, while higher quality wines generally had higher levels of 
citric acid. Volatile acidity and pH had an inverse relationship with quality.


### Did you observe any interesting relationships between the other features 
### (not the main feature(s) of interest)?

> Citric acid had an inverse correlation with pH which makes sense due to the 
increase in acid. Alcohol and residual sugar had very little, if any,
correlation which was suprising due to their relationship in the fermentation 
process.

### What was the strongest relationship you found?

>Alcohol was a definative relationship where the median high quality wines had 
two percent more alcohol than a low quality wine. The high quality wines had an 
alcohol content on par with most common red wines, around 12%. The lowest quality 
wines had a median alcohol content below 10%, which is much lower than other 
typical red wines. This could be due to a shorter fermentation time or less 
sugar to ferment.  

# Multivariate Plots Section

```{r echo=FALSE, Multivariate_Plots}

#drop_col2 <- c('X')
corr_df <- df
#corr_df <- corr_df[ , !(names(corr_df) %in% drop_col)]
corr_df[['quality']] <- as.integer(df[['quality']])

corr <- round(cor(corr_df), 1)

ggcorrplot(corr, hc.order = FALSE, 
           #type = "lower", 
           lab = TRUE, 
           lab_size = 3, 
           method="circle", 
           colors = c("darkred", "white", "green2"), 
           title="Correlogram of Wine Quality")

```

> I am starting off with a correlogram using the default pearson correlation 
coefficients, which  reinforces some of the previous observations. Alcohol has 
the strongest positive correlation with quality folowed by sulphates. Volatile 
acidity had the strongest negative correlation. Citric acid has a strong 
negative correlation with pH. Density has several strong correlations worth
exploring.

```{r echo = FALSE}

g <- ggplot(df, aes(log10(density), (citric.acid)))
g + geom_jitter(aes(  stroke = 2,  alpha = 0.5, color = factor(quality)))+
geom_smooth(aes(col=factor(quality)), method="loess", se=F)+
scale_color_brewer(type = "div", palette = "RdBu")

```

>Citric acid is one of the features from the bivariate section that correlated 
with quality. This chart shows citric acid charted with density. Density had 
several strong correlations in the correlogram including those with citric acid 
and to a lesser extent quality. This chart shows higher quality wines generally 
have higher citric acid and a lower density. 

```{r echo = FALSE}
g <- ggplot(df, aes(log10(density), log10(alcohol)))
g + geom_jitter(aes(  stroke = 2,  alpha = 0.7, color = factor(quality)))+
geom_smooth(aes(col=factor(quality)), method="loess", se=F)+
scale_color_brewer(type = "div", palette = "RdBu")
```

>After looking at this chart I realized that a high alcohol content would 
necessarily lead to a lower density due to the relatively low density of 
alcohol. There is a very distinct trend of high quality wines in the second 
quadrant and lower quality wines in the fourth quadrant. This collaborates 
previous evidence of high quality wines having higher alcohol content and lower 
density.

```{r echo = FALSE}

g <- ggplot(df, aes(log10(volatile.acidity), log10(alcohol)))
g + geom_jitter(aes(  stroke = 2,  alpha = 0.7, color = factor(quality)))+
geom_smooth(aes(col=factor(quality)), method="loess", se=F)+
scale_color_brewer(type = "div", palette = "RdBu")

```

> Volatile acidity and alcohol had the highest correlation to quality in the 
correlogram, which supported the evidence from the bivariate plots. This 
relationship can be seen here. The weak relationship between alcohol and 
volatile acidity is not evident in this graph.   

```{r echo = FALSE}
g <- ggplot(df, aes(citric.acid, log10(pH)))
g + geom_jitter(aes(  stroke = 2,  alpha = 0.7, color = factor(quality)))+
geom_smooth(aes(col=factor(quality)), method="lm", se=F)+
scale_color_brewer(type = "div", palette = "RdBu")

```

>A strong negative correlation between citric acid and pH was demonstrated in 
the bivatiate section, here we see this relationship and the correlation to
quality. Most high quality wines are high in citric acid and of lower pH. 

```{r echo = FALSE}
g <- ggplot(df, aes(log10(volatile.acidity), (citric.acid)))
g + geom_jitter(aes(  stroke = 2,  alpha = 0.7, color = factor(quality)))+
geom_smooth(aes(col=factor(quality)), method="loess", se=F)+
scale_color_brewer(type = "div", palette = "RdBu")
```

>Volatile acidity and citric acid are two of the features from the bivariate
section and the correlogram that showed a strong corrrelation to quality. This
graph shows the lowest quality wines grouped in the fourth quadrant with
very low or zero citric acid and high volatile acidity.




# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the 
### investigation. Were there features that strengthened each other in terms of
### looking at your feature(s) of interest?

>The pearson correlation coefficients showed the strongest correlations with 
quality were from alcohol, citric acid, and sulphates on the positive side and 
volitile acidity on the negative side. Alcohol showed had the highest Pearson 
coefficient with a value of +0.5. Density and alcohol were plotted together due 
to the negative correlation and a relationship between low density, high alcohol
and high quality was seen. The plot of citric acid and density also shows the 
low density wines being of higher quality.


### Were there any interesting or surprising interactions between features?

> Citric acid and volitile acidity showed a very distinct relationship of high
citric acid and low volatile acidity with the quality of wine.

### OPTIONAL: Did you create any models with your dataset? Discuss the 
### strengths and limitations of your model.

> No models were created

------

# Final Plots and Summary

### Plot One
```{r echo = FALSE}
g <- ggplot(df, aes(citric.acid))
g + geom_density(aes(fill=factor(quality)), alpha=0.6) + 
    labs(title="Citric Acid Density Grouped by Quality", 
         x="Citric Acid g/dm³",
         fill=factor('Quality'))
```

### Description One
>The density plot of citric acid was interesting because it showed the 
correlation to quality based on the density of the observations which allowed 
for a balanced perspective of the groups with very few observations.


### Plot Two
```{r echo = FALSE}

g <- ggplot(df, aes(log10(density), log10(fixed.acidity)))
g + geom_jitter(aes(  stroke = 2,  alpha = 0.5, color = factor(quality)))+
geom_smooth(aes(col=factor(quality)), method="loess", se=F)+
scale_color_brewer(type = "div", palette = "RdBu")+
labs(title="Fixed Acidity and Density Grouped by Quality", 
         subtitle=" log10 of density and fixed.acidity",
         x="Density g/cm³ (log10)",
         y= "Fixed Acidity g/dm³ (log10)",
         color=factor('Quality'))
```

### Description Two

>The near linear relationship between two seemingly unrelated features made this 
an interesting plot. It also shows the relationship of quality to these features
maintains a near linear relationship.

### Plot Three
```{r echo = FALSE}
g <- ggplot(df, aes(log10(volatile.acidity), (citric.acid)))
g + geom_jitter(aes(  stroke = 2,  alpha = 0.7, color = factor(quality)))+
geom_smooth(aes(col=factor(quality)), method="loess", se=F)+
scale_color_brewer(type = "div", palette = "RdBu")+
labs(title="Volatile Acidity and Citric AcidGrouped by Quality", 
         subtitle=" log10 of Volatile Acidity",
         x="Volatile Acidity g/dm³ (log10)",
         y= "Citric Acid g/dm³",
         color=factor('Quality'))
```

### Description Three

> This chart shows nearly all low quality observations grouped in quadrant IV
with higher quality wines near quadrand II. I think these variables alone could
produce a resonable estimation of the quality of wine.

------

## Reflection

>The red wine dataset was an interesting dataset of physicochemical and sensory 
data for the red variant of Portuguese "Vinho Verde" wine. I was able to find 
some interesting correlations between the various features and quality. Alcohol
, volatile acidity, and citric acid were all correlated to wine quality.I think 
one of the more difficlut things in this dataset is the narrow IQR of 5-6, for 
quality. There were very few observations on either extreme which made 
visualizing counts by color difficult. I was able to overcome this by using 
density plots in some situations and by using muted colors for the more common
values on other plots. Two things I found suprising in this dataset were the 
large variance of alcohol, and the lack of citric acid in some samples. I think 
more samples on the extremes of quality would add a lot to the data. As is the 
dataset does not contain any samples of 1, 2, 9, or 10, and very few samples 
for 3, 4, and 8. I beleive modeling the data using a support vector machine as 
mentioned in the wine quality info text would add to the analysis.

## Resources
* <http://r-statistics.co/Top50-Ggplot2-Visualizations-MasterList-R-Code.html>
* <https://stackoverflow.com>
* <https://drsimonj.svbtle.com/quick-plot-of-all-variables>
* <http://www.sthda.com/english/wiki/ggcorrplot-visualization-of-a-correlation-matrix-using-ggplot2>
* <https://rstudio.com/wp-content/uploads/2016/03/rmarkdown-cheatsheet-2.0.pdf>
* <https://bookdown.org/yihui/rmarkdown/html-document.html>

## References
 P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis. 
 Modeling wine preferences by data mining from physicochemical properties.
 In Decision Support Systems, Elsevier, 47(4):547-553. ISSN: 0167-9236.
